### 2024-07-03 basic-statistics.qmd
20240703 - conclusion section

If the *p*-value is less than .05, we reject the null hypothesis ($H_0$) that there was no difference or association.
Thus, we conclude that there was a statistically significant (non-zero) difference or association.
If the *p*-value is greater than .05, we fail to reject the null hypothesis; the difference/association was not statistically significant.
Thus, we do not have confidence that there was a difference or association.
However, we do not accept the null hypothesis; it could be there we did not observe an effect because we did not have adequate power to detect the effect—e.g., if the effect size was small, the data were noisy, and the sample size was small and/or unrepresentative.

@@

- the sample size (*N*): the larger the *N*, the greater the power
    - for group comparisons, the power depends on the sample size of each group
- the [effect size](#sec-practicalSignificance): the larger the effect, the greater the power
    - for group comparisons, larger effect sizes reflect:
        - larger between-group variance, and
        - smaller within-group variance (i.e., strong measurement precision, i.e., [reliability](#sec-reliability))
- the alpha level: the researcher specifies the alpha level (though it is typically set at .05); the higher the alpha level, the greater the power; however, the higher we set the alpha level, the higher the likelihood of Type I errors (false positives)
- one- versus two-tailed tests: one-tailed tests have higher power than two-tailed tests
- [within-subject](#sec-withinSubject) versus [between-subject](#sec-betweenSubject) comparisons: [within-subject designs](#sec-withinSubject) tend to have greater power than [between-subject designs](#sec-betweenSubject)

@@

Statistical significance involves the *consistency* of an effect/association/difference; it suggests that the association/difference is reliably non-zero.
However, just because something is statistically significant does not mean that it is important.
For instance, consider that we discover that players who consume sports drink before a game tend to perform better than players who do not ($p < .05$).
However, what if consumption of sports drinks is associated with an average improvement of 0.002 points per game.
A small effect such as this might be detectable with a large sample size.
This effect would be considered to be reliable/consistent because it is statistically significant.
However, such an effect would not be practically important because it has an effect size so small that it results in differences that are not practically important.
Thus, in addition to statistical significance, it is also important to consider practical significance.

@@

- [effect size](#sec-practicalSignificance)
- sample size
- [alpha level](#sec-statisticalSignificance)

Knowing any three of the following, you can calculate the fourth: statistical power, [effect size](#sec-practicalSignificance), sample size, and [alpha level](#sec-statisticalSignificance).
Below is `R` code for calculating power for each of various statistical tests (i.e., a *power analysis*).
For free point-and-click software for calculating statistical power, see G\*Power: <https://www.psychologie.hhu.de/arbeitsgruppen/allgemeine-psychologie-und-arbeitspsychologie/gpower.html>

@@

@@

There are free tools for calculating power for tests of [mediation](#mediation) and [moderation](#moderation):

@@

>>>>

If the *p*-value is less than .05, we reject the null hypothesis ($H_0$) that there was no difference or association.
Thus, we conclude that there was a statistically significant (non-zero) difference or association.
If the *p*-value is greater than .05, we fail to reject the null hypothesis; the difference/association was not statistically significant.
Thus, we do not have confidence that there was a difference or association.
However, we do not accept the null hypothesis; it could be there we did not observe an effect because we did not have adequate power to detect the effect—e.g., if the [effect size](#sec-practicalSignificance) was small, the data were noisy, and the [sample size](#sec-sampleVsPopulation) was small and/or unrepresentative.

@@

- the [sample size](#sec-sampleVsPopulation) (*N*): the larger the *N*, the greater the power
    - for group comparisons, the power depends on the [sample size](#sec-sampleVsPopulation) of each group
- the [effect size](#sec-practicalSignificance): the larger the effect, the greater the power
    - for group comparisons, larger effect sizes reflect:
        - larger between-group variance, and
        - smaller within-group variance (i.e., strong measurement precision, i.e., [reliability](#sec-reliability))
- the alpha level: the researcher specifies the alpha level (though it is typically set at .05); the higher the alpha level, the greater the power; however, the higher we set the alpha level, the higher the likelihood of Type I errors (false positives)
- one- versus two-tailed tests: one-tailed tests have higher power than two-tailed tests
- [within-subject](#sec-withinSubject) versus [between-subject](#sec-betweenSubject) comparisons: [within-subject designs](#sec-withinSubject) tend to have greater power than [between-subject designs](#sec-betweenSubject)

@@

Statistical significance involves the *consistency* of an effect/association/difference; it suggests that the association/difference is reliably non-zero.
However, just because something is statistically significant does not mean that it is important.
For instance, consider that we discover that players who consume sports drink before a game tend to perform better than players who do not ($p < .05$).
However, what if consumption of sports drinks is associated with an average improvement of 0.002 points per game.
A small effect such as this might be detectable with a large [sample size](#sec-sampleVsPopulation).
This effect would be considered to be reliable/consistent because it is statistically significant.
However, such an effect is so small that it results in differences that are not [practically important](#sec-practicalSignificance).
Thus, in addition to statistical significance, it is also important to consider [practical significance](#sec-practicalSignificance).

@@

- [effect size](#sec-practicalSignificance)
- [sample size](#sec-sampleVsPopulation)
- [alpha level](#sec-statisticalSignificance)

Knowing any three of the following, you can calculate the fourth: statistical power, [effect size](#sec-practicalSignificance), [sample size](#sec-sampleVsPopulation), and [alpha level](#sec-statisticalSignificance).
Below is `R` code for calculating power for each of various statistical tests (i.e., a *power analysis*).
For free point-and-click software for calculating statistical power, see G\*Power: <https://www.psychologie.hhu.de/arbeitsgruppen/allgemeine-psychologie-und-arbeitspsychologie/gpower.html>

@@

When designing a study, it is important to consider power and the [sample size](#sec-sampleVsPopulation) needed to detect the hypothesized effect size.
If your [sample size](#sec-sampleVsPopulation) is too small and you do not detect an effect (i.e., $p > .05$), you do not know whether your failure to detect the effect was because a) the effect does not exist, or b) the effect exists but you did not have enough power to detect it.

@@

There are free tools for calculating power for tests of [mediation](#sec-mediation) and [moderation](#sec-moderation):

@@

## Conclusion {#sec-basicStatsConclusion}

[Descriptive statistics](#sec-descriptiveStatistics) are used to describe the data, including the center, spread, or shape of data.
[Inferential statistics](#sec-inferentialStatistics) are used to draw inferences regarding differences between groups or associations between variables.
Null hypothesis signficance testing is a framework for [inferential statistics](#sec-inferentialStatistics), in which there is a [null hypothesis](#sec-nullHypothesis) and [alternative hypothesis](##sec-alternativeHypothesis).
The null hypothesis is that there is no difference between groups or that there is no association between variables.
[Statistical significance](#sec-statisticalSignificance) is evaluated with a $p$-value, which represents the probability of obtaining a result at least as extreme as the result observed if the null hypothesis is true.
Effects with *p*-values less than .05 are considered statistically significant.
However, it is also important to consider [practical significance](#sec-practicalSignificance) and effect sizes.
When designing a study, it is important to consider [statistical power](#sec-statisticalPower) and the [sample size](#sec-sampleVsPopulation) needed to detect the hypothesized effect size.
You can determine a study's [power](#sec-statisticalPower) based on the [effect size](#sec-practicalSignificance), [sample size](#sec-sampleVsPopulation), and [alpha level](#sec-statisticalSignificance).

### 2024-07-03 causal-inference.qmd
20240703 - conclusion section

>>>>

## Conclusion {#sec-causalInferenceConclusion}

There are three criteria for establishing causality: 1) the cause precedes the effect.
2) The cause is related to the effect.
3) There are no other alternative explanations for the effect apart from the cause.
Various experimental and quasi-experimental designs and approaches can be leveraged to more closely approximate causal inferences.
Longitudinal designs, within-subject analyses, inclusion of control variables, and genetically informed designs are all quasi-experimental designs that afford the researcher greater control over some possible third variable [confounds](#sec-causalDiagramConfounding).
Causal diagrams can be a useful tool for identifying the proper variables to control for (and those not to control for).
When [confounding](#sec-causalDiagramConfounding) exists, it is important to control for the [confound(s)](#sec-causalDiagramConfounding).
It is important not to control for mediators when interested in the total effect of the [predictor variable](#sec-correlationalStudy) on the [outcome variable](#sec-correlationalStudy).
When there is a [collision](#sec-colliderBias), it is important not to control for the [collider](#sec-colliderBias) unless the [collider](#sec-colliderBias) is also a [confound](#sec-causalDiagramConfounding).

### 2024-07-03 correlation.qmd
20240703 - conclusion section

Correlation is a metric of the association between variables.
Covariance is the association between variables and in an unstandardized metric that differs for variables with different scales.
By contrast, correlation is in a standarized metric that does not differ for variables with different scales.
When examining the association between variables that are [interval](#sec-interval) or [ratio](#sec-ratio) levels of measurement, Pearson correlation is used.
When examining the association between variables that are [ordinal](#sec-ordinal) in level of measurement, Spearman correlation is used.
Pearson correlation is an index of the *linear* association between variables.
If a nonlinear association is present, other indices like xi [$\xi$; @Chatterjee2021] and distance correlation coefficients are better suited to detect the association.

@@

>>>>

Correlation is an index of the association between variables.
Covariance is the association between variables and in an unstandardized metric that differs for variables with different scales.
By contrast, correlation is in a standarized metric that does not differ for variables with different scales.
When examining the association between variables that are [interval](#sec-interval) or [ratio](#sec-ratio) levels of measurement, Pearson correlation is used.
When examining the association between variables that are [ordinal](#sec-ordinal) in level of measurement, Spearman correlation is used.
Pearson correlation is an index of the *linear* association between variables.
If a nonlinear association is present, other indices like xi [$\xi$; @Chatterjee2021] and distance correlation coefficients are better suited to detect the association.

@@

## Correlation Does Not Imply Causation

As described in @sec-correlationCausation, correlation does not imply causation.
There are several reasons (described in @sec-correlationCausation) that, just because `X` is correlated with `Y` does not necessarily mean that `X` causes `Y`.
However, correlation can still be useful.
In order for two processes to be causally related, they must be associated.
That is, association is necessary but insufficient for causality.

## Conclusion {#sec-correlationConclusion}

Correlation is an index of the association between variables.
The correlation coefficient ($r$) ranges from −1 to +1, and indicates the sign and magnitude of the association.
Although correlation does not imply causality, identifying associations between variables can still be useful because association is a necessary (but insufficient) condition for causality.

### 2024-07-03 multiple-regression.qmd
20240703 - conclusion section

>>>>

## Conclusion {#sec-multipleRegressionConclusion}

Multiple regression allows examining the association between multiple [predictor variables](#sec-correlationalStudy) and one [outcome variable](#sec-correlationalStudy).
Inclusion of multiple predictors in the model allows for potentially greater predictive accuracy and identification of the extent to which each variable uniquely contributes to the outcome variable.
As with [correlation](#sec-correlation), an association does not imply causation.
However, identifiying associations is important because associations are a necessary (but insufficient) condition for causality.
When developing a multiple regression model, it is important to pay attention for potential [multicollinearity](#sec-multipleRegressionMulticollinearity)—it may become difficult to detect a given [predictor variable](#sec-correlationalStudy) as [statistically significant](#sec-statisticalSignificance) due to the greater uncertainty around the parameter estimates.

### 2024-07-03 research-methods.qmd
20240703 - conclusion section

In research, it is important to distinguish between the sample and the target population.
The target *population* is who you want your study's findings to generalize to.
For instance, if we want our findings to lead to inferences we can draw regarding all current NFL players, then NFL players are our target population.
However, despite our best efforts to recruit all NFL players into our study, we may not succeed in doing that.
The participants (i.e., people or players) who we successfully recruit to be in our study represent our *sample*.

@@

A mnemonic to help remember the difference between [mediation](#mediation) and [moderation](#moderation) is in @fig-mediationModerationMnemonic.

@@

### Mediation {#mediation}

#### Overview {#mediationOverview}

@@

#### Types of Mediation {#mediationTypes}

##### Full Mediation {#fullMediation}

@@

##### Partial Mediation {#partialMediation}

@@

##### Multiple Mediators {#multipleMediators}

@@

### Moderation (i.e., Interaction) {#moderation}

#### Overview {#moderationOverview}

@@

>>>>

In research, it is important to distinguish between the sample and the target population.
The target *population* is who you want your study's findings to generalize to.
For instance, if we want our findings to lead to inferences we can draw regarding all current NFL players, then NFL players are our target population.
However, despite our best efforts to recruit all NFL players into our study, we may not succeed in doing that.
The participants (i.e., people or players) who we successfully recruit to be in our study represent our *sample*.
The number of participants in the study is our *sample size*.

@@

A mnemonic to help remember the difference between [mediation](#sec-mediation) and [moderation](#sec-moderation) is in @fig-mediationModerationMnemonic.

@@

### Mediation {#sec-mediation}

#### Overview {#sec-mediationOverview}

@@

#### Types of Mediation {#sec-mediationTypes}

##### Full Mediation {#sec-fullMediation}

@@

##### Partial Mediation {#sec-partialMediation}

@@

##### Multiple Mediators {#sec-multipleMediators}

@@

### Moderation (i.e., Interaction) {#sec-moderation}

#### Overview {#sec-moderationOverview}

@@

## Conclusion {#sec-researchMethodsConclusion}

There are various types of research designs.
Each type of research design differs in the extent to which it supports the ability to draw causal inferences ([internal validity](#sec-internalValidity)) versus the extent to which it supports the ability to identify processes that generalize to the real-world ([external validity](#sec-externalValidity)).
In addition, it is important to understand the distinction between [sample](#sec-sampleVsPopulation) and [population](#sec-sampleVsPopulation), and the distinction between [mediation](#sec-mediation) and [moderation](#sec-moderation).
It is also important to consider the [levels of measurement](#sec-levelsOfMeasurement) used because they constrain the types of analyses that may be performed.
In addition, it is important to consider the [psychometrics](#sec-psychometrics) of measurements, including multiple aspects of [reliability](#sec-reliability) (consistency) and [validity](#sec-validity) (accuracy).
